---
title: "Lake Bilancino EDA"
author: "Alex Federspiel"
date: "3/11/2021"
output: pdf_document
---

```{r packages, include=FALSE}
  library(readr)
  library(tidyverse)
  library(lubridate)
  library(lattice)
```

```{r import data, include=FALSE}
  ## I think you'll probably have to change this path.
  Lake_Bilancino <- read_csv("C:/Users/afede/Desktop/Grad School/Data Science/Project/Alex_Federspiel_Project/Lake_Bilancino.csv")
```

## The Data
This paper will be focusing on the Lake Bilancino dataset supplied by the Acea
Group. The dataset is a collection of 6603 observations made from 2002 to 2020.
It contains the measure of rainfall in five different cities surrounding Lake
Bilancino, as well as the the temperature recordings from one of the cities. The
lake itself has its water level and flow rate into it measured.

The Acea Group is an Italian multiutily operator which manages and develops 
water and electricity networks. The water sector of their company supplied this
data with the goal of efficiently supplying their consumers with water.

## Variables
* Date
  + Uniquely identifies a day.
  + The data is ordinal and originally in the dd/mm/yyyy format. It spans from
  3/06/2002 to 30/06/2020
  + This is a potential predictor.
* Rainfall_**X**
  + The measure of rainfall in city **X** in millimeters.
  + The data is nonnegative and continuous. The units are in mm.
  + 
  
  **City**|**Min**|**1Q**|**Median**|**Mean**|**3Q**|**Max**
    ----|---|--|------|----|--|---
    **San Piero**|0|0|0|2.67|1.2|120.2
    **Mangona**|0|0|0|3.34|1.4|110
    **Sant'Agata**|0|0|0|2.67|1.2|120.2
    **Cavallina**|0|0|0|2.675|0.6|113.6
    **Le Croci**|0|0|0|3.13|1.2|88.4
  + These are potential predictors.
* Temperasure_Le_Croci
  + The Celsius temperature detected by the thermometric station in Le Croci.
  + The data is continuous. The units are in degrees Celsius.
  + 
  
  **Min**|**1Q**|**Median**|**Mean**|**3Q**|**Max**|**Range**
  ---|--|------|----|--|---|---
  -5.35|9|14.5|14.53|20.1|34|39.35
  + This is a potential predictor.
* Lake_Level
  + The lake's level measured in meters above the sea level. It has a depth of
  31 meters.
  + The data is nonnegative and continuous. The units are in m.
  + 
  
   **Measure**|**Min**|**1Q**|**Median**|**Mean**|**3Q**|**Max**|**Range**
  ---|---|--|------|----|--|---|---
  **Sea Level**|243.5|248|250.2|249.6|251.4|252.8|9.3
  **Non SL**|22.5|27|29.2|28.6|30.4|31.8|9.3
  + This is an outcome.
* Flow_Rate
  + The lake's flow rate in cubic meters per second.
  + The data is nonnegative and continuous. The units are in $m^3/s$.
  + 
  
  **Min**|**1Q**|**Median**|**Mean**|**3Q**|**Max**|**Range**
  ---|--|------|----|--|---|---
  0.31|0.6|1.5|2.76|3|74.65|74.34
  + This is an outcome.
  
## CoNVO Statement
As mentioned previously, the Acea Group is an Italian multiutility operator. 
Their water sector alone supplies upwards of nine million people in Lazio, 
Tuscany, Umbria, Molise, and Campania. With that many customers, and over that 
much area, efficiently supplying them with water can be a challenge. 

Everyone needs water. Always. But its not always as simple to say where they’ll 
get it from. The water level in waterbodies fluctuate seasonally. This means in 
the fall and winter they’ll rise and in spring and summer, they’ll drop. The 
Acea Group needs to be able to supply their customers while still preserving 
water when levels get low.

This project will use approximately 6 years of data on rainfall, temperature, 
and water levels, to build a predictive model that would forecast a waterbody’s 
level for any period. If water levels are forecasted to be low for a long 
period, the Acea Group can take the necessary steps required to start preserving
water early.

If the water sector of the Acea Group is happy with the results of the model, 
they can use it weekly to forecast the waterbody levels. It will also most 
likely need to be retrained with new additional data yearly to keep it accurate.
In five years’ time, if there were no times the water levels got dangerously 
low, this model would be a success.

## Missing Values
Looking at the first couple rows at the dataset, there seem to be a lot of 
missing values.  
```{r head, include = FALSE}
   head(Lake_Bilancino)
```
Further inspection revealed missing data for the first two years of the dataset. 
The dataset was then reduced down to only include entries from January 2 2004 to
June 30 2020. This brought it from 6603 observations down to 6025.
```{r missing values, include = FALSE}
  total_nan <- function(x) sum(is.na(x))
  Lake_Bilancino %>% summarise(across(everything(), total_nan))
  Lake_Bilancino %>% filter(is.na(Rainfall_Mangona))
  ## It seems they didn't record data until 2004 (row 578).
  Lake_Bilancino[578:nrow(Lake_Bilancino),] %>% 
    summarise(across(everything(), total_nan))
  Lake_Bilancino[578:nrow(Lake_Bilancino),] %>% 
    filter(is.na(Temperature_Le_Croci))
  ## Only remaining missing value is first temperature reading for Jan 1 04.
  ## Only going to focus on January 2nd 2004 and on for the dataset.
  Lake_Bilancino_restricted <- Lake_Bilancino[579:nrow(Lake_Bilancino),]
```

```{r turn Date into date type, include = FALSE}
  ## uses Lubridate package to turn Date into a Date-type for easier plotting
  Lake_Bilancino_restricted <- Lake_Bilancino_restricted %>% 
    mutate(Date = dmy(Date))
```

## Rainfall Delay
One challenge with forecasting water levels is that the rainfall recorded on a
particular day doesn't affect the lake on that day. It takes time for the rain
to soak into the ground and run to the river. The problem is that the exact
amount of time is unknown.

For this exploration, a variable $k$ was set up to adjust for this delay. A new
dataset is created where only the rainfall observations are shifted down in the 
dataset by $k$ days. This essentially ends up cutting off the first $k$ rows of
non-rainfall data, and the last $k$ rows of rainfall data.

With this, $k$ can easily be adjusted in once place to affect all graphs. 
Similarly, specific $k$ values can be made into datasets in needed. For the
majority of this paper, a $k$ value of $3$ will be used, but this isn't 
necessarily the correct $k$ value.

```{r building a rainfall delay dataset, include = FALSE}
  ## how many days you want to adjust the rainfall by
  k = 3

  Lake_Bilancino_adjust_k <- 
    Lake_Bilancino_restricted[(1+k):nrow(Lake_Bilancino_restricted),]
  
  for (i in 2:6) {
    Lake_Bilancino_adjust_k[,i] <- 
      Lake_Bilancino_restricted[1:(nrow(Lake_Bilancino_restricted)-k), i]
  }
```

```{r small sample data, include = FALSE}
  set.seed(42)
  mini <- floor(0.5 * nrow(Lake_Bilancino_adjust_k))
  mini_index <- sample(1:nrow(Lake_Bilancino_adjust_k), mini)
```

## Pairs Plot
```{r pairs plot, echo = FALSE}
  pairs(Lake_Bilancino_adjust_k[mini_index,])
```

The pairs plot for the data set doesn't seem to give much insight into any
correlations between variables. Even when taking a subsample of the data, it
appears as a jumbled mess. The outcome variables **Lake_Level** and 
**Flow_Rate** seem to have no correlation with the predictors. One thing of note 
is the shape of **Lake_Level** under **Date** and **Temperature** under **Date**.
Unfortunately, in this diagram you can't see it, but they both have a similar
sinusoidal pattern. Because of this, **Lake_Level** was chosen as the outcome
variable. For predictors, **Temperature_Le_Croci**, **Rainfall_Le_Croci**, 
**Rainfall_S_Piero**, and **Date** were all chosen. 

## Graphical Exploration
### Boxplot
```{r rainfall boxplots, echo = FALSE}
  ggplot(Lake_Bilancino_restricted, aes(Date, Rainfall_Le_Croci)) +
    geom_col(color = "dodgerblue4") + 
    labs(title = "Rainfall in Le Croci from 2004 to 2020", x = "Date", 
         y = "Rainfall (mm)") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.8)) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```
A simple boxplot of the rainfall for Le Croci was made. Since lake level isn't 
represented, the original dataset is used, i.e. $k=0$. While the rainfall data
is different, the overall shape and summary is the same so only one plot is 
needed.

From the summary above, this data contains a lot of 0 values. Further 
transformation of these values will be required.

### Scatterplot
```{r scatterplot rainfall v lake level, echo = FALSE}
  ggplot(Lake_Bilancino_adjust_k[mini_index,], aes(x = Rainfall_S_Piero, 
                                      y = (Lake_Level - 221))) +
    geom_point(alpha = 0.2, color = "dodgerblue4") +
    labs(title = "Rainfall in San Piero vs. Lake Bilancino's Water Level",
         subtitle = "Rainfall affects Lake Level after 3 days.",
         caption = "k = 3",
         x = "Rainfall in San Piero (mm)", 
         y = "Lake Bilancino's Level (m)")
```

A closer look at scatterplot between rainfall and lake level shows there's no
obvious correlation. A subsample was taken to reduce points, and a $k$ value of
$3$ is used, but it appears that rainfall has no real affect on the lake's water
level. The plethora of zero values for the rainfall can be easily seen here by
the dark blue line on the $y$-axis.

### Histogram
```{r histogram, echo = FALSE}
  ggplot(Lake_Bilancino_restricted, aes(factor(year(Date)), (Lake_Level - 221))) +
    geom_boxplot(color = "white", fill = "dodgerblue4") + 
    labs(title = "Histogram of Lake Bilancino's Level each Year", x = "Year", 
         y = "Lake Bilancino's Level (m)") + 
    theme_dark() +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.8))
```
This graph shows a histogram of the lake's water level at each year. Some 
noteworthy years include 2011 due to its high variance, 2012 for its low levels,
and 2014 for the outliers where apparently the lake rose above its max level of
31 m. 

### Line Graph
```{r line graph, echo = FALSE}
  ggplot(Lake_Bilancino_restricted) +
    geom_line(aes(x = Date, y = (Temperature_Le_Croci/2.5 + 22)), color = "gray") +
    geom_line(aes(Date, (Lake_Level - 221)), size = 0.65) +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.8)) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(title = "Lake Bilancino's Water Level over Time", x = "Date", 
         y = "Lake Bilancino's Level (m)")
```
As mentioned in the pairs plot section, both **Lake_Level** and **Temperature** 
had a similar shape when plotted over time. This graph plots the lake level (in
black) over time. Behind it (in gray) is a scaled down and shifted up graph of
temperature (so the $y$-values don't correspond to the actual temperature). This 
is only done highlight the similarity in patterns.

Unfortunately, this trend probably isn't due to any causation. We know that the
lake level fluctuates seasonally (rises in winter, lowers in summer) and so does 
temperature. But so far, this is the strongest connection we've seen.

### Stat Summary Plot
```{r stat summary plot, echo = FALSE}
  ggplot(Lake_Bilancino_adjust_k, 
         aes(x = Temperature_Le_Croci, y = Rainfall_Le_Croci, 
             z = (Lake_Level - 221))) +
    stat_summary_2d(bins = 10) +
    labs(title = "Temperature vs. Rainfall vs. Lake Level", 
         subtitle = "Rainfall affects Lake Level after 3 days.",
         caption = "k = 3",
         x = "Temperature in Le Croci (°C)", 
         y = "Rainfall in Le Croci (mm)", 
         fill = "Lake Water Level (m)")
```
Here, rainfall, temperature, and lake level are all plotted together. The 
number of bins was reduced to 10 in order to get a very general picture. If we
look at the boxes where water level is the lowest (potential drought territory)
we see that they're when rainfall is above 50 mm and the temperature is below 
$45^{\circ}\text{F}$ or above $59^{\circ}\text{F}$. Whereas rainfall above 50 mm
and within that temperature range seems to relate to an almost full lake. 

## Single Plot
```{r single plot over time, echo = FALSE}
  ggplot(Lake_Bilancino_adjust_k[mini_index,]) +
    theme_dark() +
    geom_col(aes(Date, Rainfall_Le_Croci), color = "#c86b85") +
    geom_col(aes(Date, Rainfall_S_Piero), color = "#1fab89") +
    geom_line(aes(Date, Temperature_Le_Croci), color = "#fdffab") +
    geom_line(aes(Date, (Lake_Level - 221)), color = "#defcf9", size = 1) +
    labs(title = "Rainfall, Temperature, and Lake Level over Time",
         subtitle = "Rainfall affects level after 3 days.",
         caption = "k = 3",
         y = "")
```
Finally, all of the chosen variables are plotted together over time. The blue 
line is the lake level, the yellow line is the temperature, the red bars are the
rainfall for Le Croci, and the green bars are the rainfall for San Piero. A 
subsample of the data was taken in hopes of decluttering the graph, but it's 
still a mess.

Let's instead take a closer look at a particular year.

```{r single plot over 2012, echo = FALSE}
  ggplot(Lake_Bilancino_adjust_k[
         year(Lake_Bilancino_adjust_k$Date) == 2012,]) +
    theme_dark() +
    geom_col(aes(Date, Rainfall_Le_Croci), color = "#c86b85") +
    geom_col(aes(Date, Rainfall_S_Piero), color = "#1fab89") +
    geom_line(aes(Date, Temperature_Le_Croci), color = "#fdffab") +
    geom_line(aes(Date, (Lake_Level - 221)), color = "#defcf9", size = 1) +
    labs(title = "Rainfall, Temperature, and Lake Level over the Year 2012",
         subtitle = "Rainfall affects level after 3 days.",
         caption = "k = 3",
         y = "")
```
The year 2012 was chosen due to its particularly low levels in the histogram 
plot. A potential relation can be seen here where rainfall relates to lake level
rising and no rainfall relates to lake level lowering. Whether or not it's a 
cause is undecided at this time.

## Causes
Unfortunately, not much can be said for causality between the variables. Both 
rainfall and temperature seem to relate to lake level but the exploration into 
the two gave no clear indication of causation.

The most likely scenario in each case is a hidden third variable. Rainfall 
causes this hidden variable which in turn causes lake level and the same for 
temperature (with a different hidden variable). 

Because of the complex nature of predicting water levels, and the many factors 
that go into it, the hidden variable might actually be multiple variables. 

(see cause image in folder)

## Difficulties
In this dataset, rainfall provided the most difficulty. Trying to adjust for the
unknown delay made things hard. I couldn't be certain if the data I was viewing
was an accurate representation. I would try different $k$ values and usually it
wouldn't make much of a difference. 

Also, due to the nature of the data, it wasn't as simple to investigate as it
has been in class. Everything has to be viewed over time to show any meaning and
from there it's hard to tell if there's any correlations.

## Other Datasets
Since the Acea Group had multiple datasets on different waterbodies, one of
those could be beneficial to combine with this one. The waterbodies are said to 
be independent of each other and have different variables but if the end goal is
to predict not just Lake Bilancino but all of the waterbody's water levels, 
obviously the other datasets will be needed.